JD notes on effects

### 2022-07-18 manuscript/body.tex
Upto MPVs

\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} central estimates for regression of $x_1$ on $y$ (outcome plots).} The horizontal dashed lines are the mean of the central estimates and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively. The grey points are the binned simulated $y$. The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other input variables at their mean and the true predictions based on the simulation parameters -- central estimate. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated one in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$. Similarly, the central estimate based on the three approaches match the truth. B: Even with the simple interaction between non-focal predictors, we see differences between predicted mean, $\bar{\hat{y}}$ and simulated mean , $\bar{y}$. Similarly, there are differences between central estimates and the truth, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}

@@

In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, \FREF{justify_plots}A. However, in the presence of interactions, even as simple as the one in \EREF{simple_inter_higher}, the estimates start to differ. In particular, \pkg{emmeans} and \pkg{effects} have similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B. To generate \FREF{justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the linear predictor variables corresponding to $x_2$ and $x_3$. On other hand, to generate \FREF{justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction linear predictor variable ($x_2x_3$). In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses linear predictor variables to compute $\over{x_2x_3}$. In the case of \FREF{justify_plots}A, averaging the input variables is equivalent to averaging the linear predictor variables.

One may be interested in the uncertainties associated with the focal predictor only, excluding other uncertainties due to other non-focal predictors -- effect plots. However, as previously mentioned, the two packages do not provide a straightforward way do achieve this. To illustrate this, we consider \EREF{simple_inter_higher_no_interaction} and use $x_2$ as the focal predictor, and generate the outcome plot shown in \FREF{justify_ci_plots}. 
%
\begin{figure}
\begin{center}
\includegraphics{justify_anchors_all.ggp.pdf}
\end{center}
\caption{{\bf Prediction and effect plots.} The description of horizontal, vertical and trend lines remain the same as above. A: The wider dashed curves corresponds to the conventional prediction curves from \pkg{emmeans} and \pkg{effects} packages, while the narrower curves crossing at the mean of the focal predictor, also referred to as the center point, represent the effect from \pkg{varpred}. For simple OLS models, the effect-styled curves cross at the center point (which is the default anchor and also referred to as center-anchored). The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal predictors. On the other hand, effect-styled curves only take into account the main effect uncertainty of the focal predictor. B: center- and zero- anchored effects. The center-anchored effects (as described above) while the zero-anchored means that the anchor is at zero value of the focal predictor. The choice of the anchor do not affect the central estimates, hence they are identical in both choices of the anchor.}
\label{fig:justify_ci_plots}
\end{figure}
%
For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal predictors, but narrower and crosses at the mean of the focal predictor in \pkg{varpred}. In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., the anchor. For simple OLS models, the point where the curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate and stable choice).

Lastly, we compare effect plots and the marginal effects \citep{leeper2017interpreting} plots implemented in package \pkg{margins} \citep{lenth2018package}. To illustrate this, we use the simulated data described in \nameref{S1_Appendix}. The results are shown in \FREF{qoi_age_pred_plot}.
%
\begin{figure}
\centering
\includegraphics{cubic_varpred_margins_pred.ggp.pdf}
\caption{{\bf Effect and marginal effect plots.} \NEW{A:  The effect plot. B: The marginal effect (ME) at representative values of the focal predictor \code{age} together with the $95\%$ confidence intervals. For simple linear models with no interactions, the ME should be constant across all the values of the focal predictor; and it would correspond to the slope of the central estimate of central estimate. However, in this example, the focal predictor is a cubic polynomial, and so, the effect is not constant but a function of the partial derivatives of the corresponding linear predictor variables.}}
\label{fig:qoi_age_pred_plot}
\end{figure}
%

>>>>

\caption{{\bf A comparison of central estimates for \pkg{emmeans}, \pkg{effects} and \pkg{varpred}.} The horizontal dashed lines are the mean of the central estimates and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively. The grey points are the binned simulated $y$. The trend lines represent the $\hat{y}$ -- central estimate. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated one in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$. Similarly, the central estimate based on the three approaches match the truth. B: Even with the simple interaction between non-focal predictors, we see differences between predicted mean, $\bar{\hat{y}}$ and simulated mean, $\bar{y}$. Similarly, there are differences between central estimates and the truth, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}

@@

In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce identical estimates, which match the simulated values, \FREF{justify_plots}A. However, in the presence of interactions, even as simple as the one in \EREF{simple_inter_higher}, the estimates start to differ. In particular, \pkg{emmeans} and \pkg{effects} have similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), \FREF{justify_plots}B. To generate \FREF{justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the linear predictor variables corresponding to $x_2$ and $x_3$. On other hand, to generate \FREF{justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction linear predictor variable ($x_2x_3$). In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses linear predictor variables to compute $\over{x_2x_3}$. In the case of \FREF{justify_plots}A, averaging the input variables is equivalent to averaging the linear predictor variables.

One may be interested in the uncertainties associated with the focal predictor only, excluding other uncertainties due to other non-focal predictors -- effect plots. However, as previously mentioned, the two packages do not provide a straightforward way do achieve this. To illustrate this, we consider \EREF{simple_inter_higher_no_interaction} and use $x_2$ as the focal predictor, and generate the outcome plot shown in \FREF{justify_ci_plots}. 
%
\begin{figure}
\begin{center}
\includegraphics{justify_anchors_all.ggp.pdf}
\end{center}
\caption{{\bf Prediction and effect plots.} The description of horizontal, vertical and trend lines remain the same as above. A: The wider dashed curves corresponds to the conventional prediction curves from \pkg{emmeans} and \pkg{effects} packages, while the narrower curves crossing at the center point represent the effect from \pkg{varpred}. For simple OLS models, the effect-styled curves cross at the center point (which is the default anchor and also referred to as center-anchored). The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal predictors. On the other hand, effect-styled curves only take into account the main effect uncertainty of the focal predictor. B: center- and zero- anchored effects. The zero-anchored effects means that the anchor is at zero value of the focal predictor. The choice of the anchor do not affect the central estimates, hence they are identical in both choices of the anchor.}
\label{fig:justify_ci_plots}
\end{figure}
%
For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal predictors, but narrower and crosses at the mean of the focal predictor in \pkg{varpred}. In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal predictor we are more certain about, i.e., the anchor. For simple OLS models, the point where the curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate and stable choice).

Lastly, we compare effect plots and the marginal effects \citep{leeper2017interpreting} plots implemented in package \pkg{margins} \citep{lenth2018package}. To illustrate this, we use the simulated data described in \nameref{S1_Appendix}. The results are shown in \FREF{qoi_age_pred_plot}.
%
\begin{figure}
\centering
\includegraphics{cubic_varpred_margins_pred.ggp.pdf}
\caption{{\bf Effect and marginal effect plots.} \NEW{A:  The effect plot. B: The marginal effect (ME) at representative values of the focal predictor \code{age} together with the $95\%$ confidence intervals. For simple linear models with no interactions, the ME should be constant across all the values of the focal predictor; and it would correspond to the slope of the central estimate. However, in this example, the focal predictor is a cubic polynomial, and so, the effect is not constant but a function of the partial derivatives of the corresponding linear predictor variables.}}
\label{fig:qoi_age_pred_plot}
\end{figure}
%
